---
title: "Plotting xcms objects with lcmsPlot"
package: xcms
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Plotting xcms objects with lcmsPlot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{xcms,MsExperiment,faahKO,BiocStyle}
  %\VignettePackage{xcms}
  %\VignetteKeywords{mass spectrometry, metabolomics, visualisation}
---

```{r biocstyle, echo = FALSE, results = "asis"}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("xcms")`<br />
**Compiled**: `r date()`

```{r init, message = FALSE, echo = FALSE, results = "hide"}
library(BiocStyle)
library(xcms)
library(MsExperiment)
library(faahKO)
register(SerialParam())
has_lcmsplot <- requireNamespace("lcmsPlot", quietly = TRUE)
knitr::opts_chunk$set(
    fig.width = 9, fig.height = 5, fig.align = "center", out.width = "100%",
    warning = FALSE, message = FALSE
)
```

# Introduction

[lcmsPlot](https://bioconductor.org/packages/release/bioc/html/lcmsPlot.html) is an R Bioconductor package designed for visualising
Liquid Chromatography-Mass Spectrometry (LC-MS) data with high-quality plots
suitable for publication. The package enables users to generate and customise chromatograms,
mass traces, spectra, and more with fine-tuned aesthetics and annotation options.

A plot begins with a data object and gains layers with `+`:

```r
lcmsPlot(data) +
    lp_chromatogram(features = ...) +
    lp_spectra(mode = "closest_apex") +
    lp_facets(facets = "feature_id")
```

Each `lp_*()` layer either adds something to draw (chromatograms, spectra, mass
traces, intensity maps, peak densities) or controls how the resulting panels are
arranged and labelled. The data object can be an `XcmsExperiment` or `XCMSnExp`,
an `MsExperiment`, a `character` vector of mzML paths, or the output of MZmine,
MS-DIAL, Compound Discoverer or LipidSearch.

There are two ways to reach it from `xcms`, and this vignette covers both:

- **Directly**, by composing a figure from lcmsPlot layers.
- **Through `xcms`**, by passing `backend = "lcmsPlot"` to a plotting method you
  already call. The default `backend = "base"` keeps the original base-R output.

`lcmsPlot` is a *Suggests*-only dependency of `xcms`, and version 1.1.8 or later
is required:

```r
BiocManager::install("lcmsPlot")
```

The development version lives on GitHub:

```r
remotes::install_github("computational-metabolomics/lcmsPlot")
```

This vignette is about where the two packages meet. For lcmsPlot itself, i.e., the
full range of plot types, interoperability with vendor formats, and publication
theming, see its own vignettes, which are not reproduced here:

```r
browseVignettes("lcmsPlot")
vignette("lcms_data_plotting", package = "lcmsPlot")
vignette("creating_publication_ready_plots", package = "lcmsPlot")
```

```{r note-no-lcmsplot, echo = FALSE, results = "asis", eval = !has_lcmsplot}
cat("> **Note:** `lcmsPlot` is not available in this session.",
    "The base-graphics panels below render normally;",
    "the lcmsPlot panels are skipped.\n")
```

# Setup

```{r setup-data, message = FALSE}
fls <- dir(system.file("cdf", package = "faahKO"),
           recursive = TRUE, full.names = TRUE)[c(1, 2, 7)]
pd  <- data.frame(sample = sub("\\.CDF$", "", basename(fls)),
                  group  = sub("([0-9]+)$", "", sub("\\.CDF$", "",
                                                    basename(fls))))
mse <- readMsExperiment(spectraFiles = fls, sampleData = pd)

## XcmsExperiment with detected peaks, grouped, RT-adjusted
cwp   <- CentWaveParam(noise = 10000, snthresh = 40,
                       prefilter = c(3, 10000))
xmse  <- findChromPeaks(mse, param = cwp)
xmseg <- groupChromPeaks(xmse,
                         param = PeakDensityParam(sampleGroups = rep(1, 3)))
xmsegr <- adjustRtime(xmseg, param = PeakGroupsParam(span = 0.4))

## XChromatograms (two m/z windows × three samples) with peaks
mzr   <- matrix(c(335, 335, 344, 344), ncol = 2, byrow = TRUE)
xchrs <- findChromPeaks(chromatogram(mse, mz = mzr),
                        param = CentWaveParam())

## plotChromPeakDensity() operates on a single EIC across samples
xchr1 <- xchrs[1, , drop = FALSE]

## A single xcmsRaw for the legacy methods
xraw <- xcmsRaw(fls[1], profstep = 1)
```

# Why plot with lcmsPlot?

The clearest way to see what the layered interface buys you is to write the
same figure twice. Because the layers compose, one plot object can hold
chromatograms, their spectra, peak highlighting and a panel layout at once, and
each of those is a single line.

The figure below is a good example: for two features in two samples, an
extracted ion chromatogram with its detected peak shaded, and directly
underneath it the MS1 spectrum recorded at that peak's apex.

```{r apex-setup}
xmse_apex <- findChromPeaks(
    mse, param = CentWaveParam(peakwidth = c(20, 80), noise = 5000,
                               prefilter = c(6, 5000)))

feats <- rbind(
    c(mzmin = 334.9, mzmax = 335.1, rtmin = 2700, rtmax = 2900),
    c(mzmin = 343.9, mzmax = 344.1, rtmin = 2600, rtmax = 2750))
samples <- c("ko15", "ko16")
```

## With lcmsPlot

```{r apex-lcms, eval = has_lcmsplot, fig.width = 10, fig.height = 9}
library(lcmsPlot)

lcmsPlot(xmse_apex, sample_id_column = "sample") +
    lp_chromatogram(features = feats, sample_ids = samples,
                    highlight_peaks = TRUE) +
    lp_spectra(mode = "closest_apex", ms_level = 1) +
    lp_facets(facets = c("feature_id", "sample_id"), ncol = 2) +
    lp_legend(position = "bottom") +
    lp_layout(design = "C\nS\nS")
```

`lp_spectra(mode = "closest_apex")` is what does the work: it reads the
retention time of each detected chromatographic peak and pulls the nearest MS1 scan. It needs no m/z or retention
time arguments of its own because it inherits the features and samples from
`lp_chromatogram()`. `lp_facets()` then splits *both* panels, since faceting
happens at the *patchwork* level rather than inside a single `ggplot`, and
`lp_layout(design = "C\nS\nS")` gives the chromatogram one row and the spectrum
two. The letters are the first letters of the dataset names.

## The same figure with base xcms

```{r apex-base, fig.width = 10, fig.height = 9}
layout(matrix(seq_len(2 * nrow(feats) * length(samples)),
              nrow = 2 * nrow(feats)))
par(mar = c(4, 4, 2, 1))

for (s in samples) {
    i <- match(s, sampleData(xmse_apex)$sample)
    sps <- filterMsLevel(spectra(xmse_apex[i]), 1L)
    for (k in seq_len(nrow(feats))) {
        chr <- chromatogram(xmse_apex[i], mz = feats[k, 1:2],
                            rt = feats[k, 3:4])
        plot(chr, main = paste(s, "- m/z", round(feats[k, 1], 3)))
        pks <- chromPeaks(chr)
        apex <- pks[which.max(pks[, "maxo"]), "rt"]
        sp <- sps[which.min(abs(rtime(sps) - apex))]
        plot(mz(sp)[[1]],
             100 * intensity(sp)[[1]] / max(intensity(sp)[[1]]),
             type = "h", xlab = "m/z", ylab = "Rel. intensity (%)",
             main = paste("apex RT", round(apex), "s"))
    }
}
```

## The difference

Apart from the lines differences (seven lines against nineteen), the base
version has to make every decision explicitly: how many panels there are and how
they are arranged (`layout()`), which sample maps to which index, how to find
the apex from the peak matrix, how to locate the nearest MS1 scan, and how to
normalise the spectrum. Each of those is a place to get it wrong, and each is
re-derived by hand every time you want a variation on the figure.

The lcmsPlot version states what should be in the figure declaratively and leaves the
arrangement to the layers. Changing it is proportionate: a third feature is one
more row in `feats`, MS2 spectra instead of MS1 is `ms_level = 2`, and spectra
across the whole peak rather than at its apex is
`mode = "across_peak", interval = 10`. In the base version each of those means
editing the loop and the layout matrix.

# Composing figures

Everything beyond the data layers is presentation.
The same extracted ion chromatograms can be coloured by an experimental factor, split into panels, annotated, or relabelled without altering how they were extracted.

```{r compose, eval = has_lcmsplot, fig.width = 10, fig.height = 5}
lcmsPlot(xmse_apex, sample_id_column = "sample") +
    lp_chromatogram(features = feats) +
    lp_arrange(group_by = "group") +
    lp_facets(facets = "feature_id", ncol = 2, free_x = TRUE) +
    lp_labels(title = "Extracted ion chromatograms by sample group",
              legend = "Group") +
    lp_legend(position = "bottom")
```

`lp_arrange()` picked the colour aesthetic off the `group` column of the sample
metadata, so the two KO samples and the WT sample are distinguished without
naming a single colour; `lp_facets()` split the two features into their own
panels and gave each its own retention time range.

| Layer | Controls |
|---|---|
| `lp_arrange(group_by)` | which metadata column becomes the colour aesthetic |
| `lp_facets(facets, ncol, nrow, free_x, free_y)` | splitting into panels by one or more metadata columns |
| `lp_grid(rows, cols, free_x, free_y)` | a panel matrix defined by a row and a column factor |
| `lp_layout(design)` | how the panel *types* stack |
| `lp_labels(title, legend)` | the plot title and the legend title |
| `lp_legend(position)` | `"top"`, `"right"`, `"bottom"`, `"left"` or `"inside"` |
| `lp_rt_line(intercept, line_type, color)` | a vertical marker at a retention time |

# Plotting xcms objects with `backend`

The rest of this vignette works through the `xcms` side: every plotting method
that accepts `backend = "lcmsPlot"`, shown next to its base-R output.

| Class | Method | Status | Notes |
|---|---|---|---|
| `XcmsExperiment` | `plotAdjustedRtime()` | ✅ Full | `lp_rt_diff_plot()` |
| `XcmsExperiment` | `plot(..., type = "XIC")` | ✅ Full | `lp_intensity_map()` + `lp_chrom_peak_rects()` |
| `XcmsExperiment` | `plotChromPeaks()` | ✅ Full | `lp_intensity_map()` + `lp_chrom_peak_rects()` |
| `XcmsExperiment` | `plotChromPeakImage()` | ✅ Full | `lp_peak_count_image()` |
| `XcmsExperiment` | `plotPrecursorIons()` | ❌ None | No lcmsPlot precursor scatter |
| `XChromatogram` | `plot()` | ✅ Full | `lp_chromatogram(highlight_peaks_mode = peakType)` |
| `XChromatograms` | `plot()` | ✅ Full | `lp_chromatogram` + `lp_facets` |
| `XChromatograms` | `plotChromPeakDensity()` | ✅ Full | `lp_chromatogram` + `lp_peak_density()` |
| `XChromatograms` | `plotChromatogramsOverlay()` | ✅ Full | `lp_chromatogram` + `lp_facets` |
| `xcmsRaw` | `plotTIC()` | ✅ Full | `lp_chromatogram(aggregation_fun = "sum")` |
| `xcmsRaw` | `plotEIC()` | ✅ Full | `lp_chromatogram(features = ...)` |
| `xcmsRaw` | `plotChrom()` | ✅ Full | `aggregation_fun` `"max"` / `"mean"` |
| `xcmsRaw` | `plotScan()` | ✅ Full | `lp_spectra(scan_index = scan)` |
| `xcmsRaw` | `plotSpec()` | ❌ None | lcmsPlot has no cross-scan spectrum averaging |
| `xcmsRaw` | `plotRaw()` | ✅ Full | `lp_intensity_map(geom = "point")` |
| `xcmsRaw` | `levelplot()` | ⚠️ Partial | Axes and `col.regions` carry across; lcmsPlot draws centroids, not the gridded profile matrix |
| `xcmsRaw` | `plotPeaks()` | ❌ None | Needs `findPeaks` matrix |
| `xcmsRaw` | `plotSurf()` | ❌ None | 3D `rgl`; lcmsPlot is `ggplot2` |
| `xcmsRaw` | `image()` | ❌ None | Out of scope |

Legend: ✅ Full — the lcmsPlot rendering reproduces the original; ⚠️ Partial —
some arguments / visual details do not carry over (see notes column and the
end-of-section bullets below); ❌ None — no lcmsPlot equivalent.

Throughout, base-graphics styling arguments (`col`, `lty`, `lwd`, `pch`,
`type`, …) and the interactive `ident` arguments are ignored on the lcmsPlot
branch: lcmsPlot applies its own scales and produces a static figure.

# XcmsExperiment

## `plotAdjustedRtime()`

```{r xmse-rtadj-base}
plotAdjustedRtime(xmsegr)
grid()
```

```{r xmse-rtadj-lcms, eval = has_lcmsplot}
plotAdjustedRtime(xmsegr, backend = "lcmsPlot")
```

*Carries across:* the requirement that alignment has been performed.
*Dropped on the lcmsPlot branch:* `col`, `lty`, `lwd`, `type`, `peakGroupsCol`,
`peakGroupsPch`, `peakGroupsLty`, `ylim` and other base-graphics styling.

## `plot(x, type = "XIC")`

```{r xmse-xic-base}
tmp <- filterMz(filterRt(xmse, rt = c(2550, 2800)), mz = c(342.5, 344.5))
plot(tmp)
```

```{r xmse-xic-lcms, eval = has_lcmsplot}
plot(tmp, backend = "lcmsPlot")
```

Both panels show the m/z–RT intensity map with the detected chromatographic
peaks overlaid as rectangles; `peakCol` sets the rectangle border on either
backend. The base panel adds the marginal chromatogram and spectrum of the
classical three-part XIC, which lcmsPlot does not draw.

## `plotChromPeaks()`

```{r xmse-chrompeaks-base}
plotChromPeaks(xmse, file = 1)
```

```{r xmse-chrompeaks-lcms, eval = has_lcmsplot}
plotChromPeaks(xmse, file = 1, backend = "lcmsPlot")
```

The base method draws the peak rectangles on an empty frame. The lcmsPlot
branch draws the same rectangles on top of the intensity map of that file, so
each peak can be read against the signal it was detected in.

Note that the faahKO files are m/z-binned, so every peak has
`mzmin == mzmax` and the rectangles collapse to segments; on profile or
centroid data they are boxes with a real m/z extent.

*Dropped on the lcmsPlot branch:* `add` (the map always starts a new panel) and
`msLevel`.

## `plotChromPeakImage()`

```{r xmse-peakimage-base}
plotChromPeakImage(xmse, binSize = 30)
```

```{r xmse-peakimage-lcms, eval = has_lcmsplot}
plotChromPeakImage(xmse, binSize = 30, backend = "lcmsPlot")
```

`binSize`, `log` and `xlim` carry across, and the bin counts match those of the
base method. Samples keep their injection order on the y-axis. *Dropped:*
`yaxt`.

# XChromatogram / XChromatograms

## `plot,XChromatogram` — single trace

```{r xchr-base}
xchr <- xchrs[[1, 1]]                   # row 1 (m/z 335), sample 1
plot(xchr)
```

```{r xchr-lcms, eval = has_lcmsplot}
plot(xchr, backend = "lcmsPlot")
```

`peakType` (`"polygon"`, `"rectangle"`, `"point"`, `"none"`) maps onto
`lcmsPlot::lp_chromatogram(highlight_peaks_mode = …)`:

```{r xchr-modes, eval = has_lcmsplot}
plot(xchr, backend = "lcmsPlot", peakType = "rectangle")
plot(xchr, backend = "lcmsPlot", peakType = "none")
```

## `plot,XChromatograms` — matrix layout

```{r xchrs-base}
plot(xchrs)
```

```{r xchrs-lcms, eval = has_lcmsplot}
plot(xchrs, backend = "lcmsPlot")
```

When the matrix has multiple rows lcmsPlot facets by the auto-generated
`feature_id`; for a single row it facets by `sample_id`.

*Dropped on the lcmsPlot branch:* `col`, `lty`, `type`, `peakCol`, `peakBg`,
`peakPch` — lcmsPlot uses its own per-sample colour scale.

## `plotChromPeakDensity()`

```{r xchr-density-base}
pdp1 <- PeakDensityParam(sampleGroups = rep(1, ncol(xchr1)))
plotChromPeakDensity(xchr1, param = pdp1)
```

```{r xchr-density-lcms, eval = has_lcmsplot}
plotChromPeakDensity(xchr1, param = pdp1, backend = "lcmsPlot")
```

Both backends stack the EICs above the peak-density panel, in which each
chromatographic peak is a point positioned by its sample. `bw`, `minFraction`,
`minSamples` and `sampleGroups` are taken from `param`, and `simulate` behaves
as it does on the base backend: `simulate = TRUE` derives the feature groups
the parameters *would* produce, `simulate = FALSE` draws the feature
definitions already stored on the object.

As with the base method, `object` has to be a single row.

## `plotChromatogramsOverlay()`

```{r xchrs-overlay-base}
plotChromatogramsOverlay(xchrs, col = c("#c0392b", "#2471a3"))
```

```{r xchrs-overlay-lcms, eval = has_lcmsplot}
plotChromatogramsOverlay(xchrs, backend = "lcmsPlot")
```

One panel per sample, with the EICs overlaid inside it. The base method
identifies the EICs through `col`, which is recycled across them if only one
colour is given — hence the explicit pair above, without which both traces are
the same grey. lcmsPlot assigns the colours itself, from the m/z window of each
EIC.

Only `XChromatograms` is supported: lcmsPlot derives the per-EIC identifiers
that the overlay colours and orders by from the m/z windows of an
`XChromatograms`, which a plain `MChromatograms` does not carry.

# Legacy xcmsRaw methods

## `plotTIC()`

```{r tic-base}
plotTIC(xraw)
```

```{r tic-lcms, eval = has_lcmsplot}
plotTIC(xraw, backend = "lcmsPlot")
```

*Dropped:* `ident`, `msident` (interactive identification).

## `plotEIC()`

```{r eic-base}
plotEIC(xraw, mzrange = c(334.9, 335.1), rtrange = c(2700, 2900))
```

```{r eic-lcms, eval = has_lcmsplot}
plotEIC(xraw, mzrange = c(334.9, 335.1), rtrange = c(2700, 2900),
        backend = "lcmsPlot")
```

*Dropped:* `add` (overlay), `scanrange`, `type`.

## `plotChrom()`

```{r chrom-base}
plotChrom(xraw, base = TRUE)
```

```{r chrom-lcms, eval = has_lcmsplot}
plotChrom(xraw, base = TRUE, backend = "lcmsPlot")
```

`base` selects the aggregation on both backends: `base = TRUE` gives the base
peak chromatogram (`aggregation_fun = "max"`) and `base = FALSE` the averaged
ion chromatogram (`aggregation_fun = "mean"`).

```{r chrom-base-false}
plotChrom(xraw, base = FALSE)
```

```{r chrom-base-false-lcms, eval = has_lcmsplot}
plotChrom(xraw, base = FALSE, backend = "lcmsPlot")
```

The mean is taken over the profile matrix, so scans with no signal in a mass
bin contribute zeros and the two traces agree exactly — provided the object was
read with a non-zero `profstep`. For file-backed inputs without a profile
matrix lcmsPlot averages over the measured peaks of each scan instead, which
sits higher.

## `plotScan()`

```{r scan-base}
plotScan(xraw, scan = 500)
```

```{r scan-lcms, eval = has_lcmsplot}
plotScan(xraw, scan = 500, backend = "lcmsPlot")
```

*Dropped:* `ident`.

## `plotSpec()`

```{r spec-base}
plotSpec(xraw)
```

`plotSpec()` averages the profile matrix across every scan in the selected
range. `lcmsPlot::lp_spectra()` extracts single scans only, so there is no
lcmsPlot equivalent and this method has no `backend` argument; passing one is
an error rather than a silently ignored argument.

For a single scan use `plotScan()`, which does have an lcmsPlot backend.

## `plotRaw()`

```{r raw-base}
plotRaw(xraw, mzrange = c(300, 350), rtrange = c(2700, 2900))
```

```{r raw-lcms, eval = has_lcmsplot}
plotRaw(xraw, mzrange = c(300, 350), rtrange = c(2700, 2900),
        backend = "lcmsPlot")
```

Both panels are a scatter of the individual centroids;
`lcmsPlot::lp_intensity_map(geom = "point")` skips the binning so gaps in the
mass traces stay visible. *Dropped:* `log`, `title` and `scanrange` — lcmsPlot
always maps intensity on a log scale.

## `levelplot()`

```{r lvl-base}
levelplot(xraw)
```

```{r lvl-lcms, eval = has_lcmsplot}
levelplot(xraw, backend = "lcmsPlot")
```

Both render a 2D intensity surface over the same m/z (x) and retention time (y)
axes, and `col.regions` is honoured as the lcmsPlot fill scale. The remaining
difference is a data model rather than a style: the *lattice* version draws the
gridded profile matrix, which has a value in every cell, so its baseline is
filled; lcmsPlot draws only the measured centroids, leaving the baseline blank
and the mass traces reading as separate lines. *Dropped:* `log` — lcmsPlot
always maps intensity on a log scale.

# Methods without an lcmsPlot equivalent

| Method | Why no lcmsPlot path | Workaround |
|---|---|---|
| `plotSpec,xcmsRaw` | `lp_spectra()` extracts single scans; it has no cross-scan averaging | Use the base backend, or `plotScan(x, scan, backend = "lcmsPlot")` for one scan |
| `plotPrecursorIons,MsExperiment` | No lcmsPlot layer plots precursor m/z against retention time | Use the base backend |
| `plotPeaks,xcmsRaw` | Consumes a `findPeaks(xcmsRaw)` matrix | Pass the peak matrix to `lcmsPlot::lp_chromatogram(features = )` and facet by `feature_id`, or migrate to `XcmsExperiment` (`findChromPeaks`) |
| `plotSurf,xcmsRaw` | 3D `rgl` surface; lcmsPlot is `ggplot2`-only | Use the base backend (requires `rgl`) |
| `image,xcmsRaw` | Redundant with `levelplot()`, and out of scope of the integration | Use the base backend, or `levelplot(x, backend = "lcmsPlot")` |

# Customising the lcmsPlot output

The lcmsPlot branches return the `patchwork`/`ggplot` object invisibly, so
you can style it after the fact:

```{r customise, eval = has_lcmsplot, fig.width = 8, fig.height = 4}
library(ggplot2)
library(patchwork)

p <- plotEIC(xraw, mzrange = c(334.9, 335.1), rtrange = c(2700, 2900),
             backend = "lcmsPlot")

p &
    theme_minimal(base_size = 11) &
    labs(title = "EIC m/z 335 - ko15")
```

To save the figure use `ggplot2::ggsave()` (or `patchwork::wrap_plots()` →
`ggsave()`):

```{r save, eval = FALSE}
ggsave("eic_ko15.pdf", p, width = 8, height = 4)
```

This is post-processing: the `backend` argument gives you no way to reach inside
the figure while it is being built. When you need that (e.g., a different title, a
legend somewhere else, panels arranged another way) build the plot from
lcmsPlot layers instead, as in *Composing figures* above.

# Session information

```{r sessioninfo}
sessionInfo()
```
